{\bf {\Large Detiding time series of dissolved oxygen for improved estimates of ecosystem metabolism}}

### 2014-08-25 assess.Rnw
complete first draft

The ability of the weighted regression to detide the \ac{DO} signal was evaluated first using a simulation approach.  Observed \ac{DO} time series were created to represent the sum of variation from biological processes and physical effects related to tidal advection:  
\begin{equation} \label{do_obs}
DO_{obs} = DO_{bio} + DO_{adv}
\end{equation}
Biological \ac{DO} signals are inherently noisy \citep{Batt12} and can be further described as:
\begin{equation} \label{do_bio} 
DO_{bio} = DO_{die} + DO_{unc}
\end{equation} 
\begin{equation} \label{do_unc}
DO_{unc} = \epsilon_{obs} + \epsilon_{proc}
\end{equation}
where the biological \ac{DO} signal is the sum of diel variation on a 24 hour scale plus uncertainty or noise.  Total uncertainty in the biological \ac{DO} signal is described as variation from observation and process uncertainty \citep{Hilborn97}.  Multiple time series at 30 minute observations over 30 days were created following \cref{do_obs,do_bio,do_unc} such that observed \ac{DO} is generalized as the additive combination of four time series (\cref{fig:do_sim}):
\begin{equation} \label{do_obs_all}
DO_{obs} = DO_{adv} + DO_{die} + \epsilon_{obs} + \epsilon_{pro}
\end{equation}
Time series were created by varying the relative magnitudes of each of the four components of observed \ac{DO} to test the effectiveness of weighted regression under different scenarios.  The effects of air-sea gas exchange were not considered in the simulation given that methods are available for \textit{in situ} data to correct observed \ac{DO} for diffusion \citep[i.e., ][]{Thebault08}.

Each parameter of the simulated time series was created as follows. First, biological \ac{DO} time series in \cref{do_bio} were created by adding noise or variance to a diel component (\cref{fig:do_sim}).  The diel component, $DO_{die}$, was estimated using a sine/cosine function \citep{Cryer08}:
\begin{equation} \label{do_sin}
DO_{die} = \alpha + \beta\cos\left(2\pi ft + \Phi\right)
\end{equation}
such that the mean DO $\alpha$ was 8, amplitude $\beta$ was 1, $f$ was 1/48 to repeat on a 24 hour period every 30 minutes, $t$ was the time series vector and $\Phi$ was the x-axis origin set for an arbitrary sunrise at 630am.  The diel signal was increasing during the day and decreasing during the night for each 24 hour period and ranged from 7 to 9 mg L$^{-1}$.  Noise or uncertainty was added to the diel \ac{DO} signal to simulate natural variation in \ac{DO} throughout the time series (\cref{fig:do_sim}).  Total uncertainty was the sum of observation and process uncertainty for $n = 1440$ (30 minutes by 30 days) observations \citep{Hilborn97}, such that:
\begin{equation} \label{do_unc_n}
DO_{unc, n} = \epsilon_{obs, n} + \int_{t = 1}^{n} \epsilon_{pro, t}
\end{equation}
where observation and process uncertainty ($\epsilon_{obs}$, $\epsilon_{pro}$) were simulated as normally distributed random variables with mean zero and standard deviation varying from zero to an upper limit, described below.  To induce auto-correlation, process uncertainty was estimated as the cumulative sum of $n$ observations where the noise at time $t+1$ was equal to the noise at time $t$ plus additional variation drawn from the normal distribution.  The noise vector for process uncertainty was rescaled to constrain the variation within the bounds for standard deviation defined by the random variable. The total uncertainty, $DO_{unc}$, was added to the diel \ac{DO} time series to create the biological \ac{DO} time series (\cref{fig:do_sim}).

A semidiurnal tidal series was simulated as a sine wave with a period of 12.5 hours to approximate the principal lunar component \citep{Foreman89}.  The amplitude was set to 1 meter and centered  at 4 meters.  Initial assessments indicated that tide type (i.e., diurnal, semidiurnal, mixed) did not signifcantly affect the outcome of the results and a semidiurnal time series was used to reduce the total number of simulations.  The tidal time series was added to the biological \ac{DO} series to simulate \ac{DO} changes with advection, $DO_{adv}$ (\cref{fig:do_sim}). Conceptually, this vector represents the rate of change in \ac{DO} as a function of horizontal water movement from tidal advection such that:
\begin{equation} \label{deltdo}
\frac{\delta DO_{adv}}{\delta t} = \frac{\delta DO}{\delta x} \cdot \frac{\delta x}{\delta t}
\end{equation}
\begin{equation} \label{deltx}
\frac{\delta x}{\delta t} = k \cdot \frac{\delta H}{\delta t}
\end{equation}
where the first derivative of the tidal time series, as change in height over time $\delta H / \delta t$, is multiplied by a constant $k$, to estimate horizontal tidal excursion over time, $\delta x / \delta t$.  The horizontal excursion is assumed to be associated with a horizontal \ac{DO} change, $\delta DO / \delta x$, such that the product of the two estimates the \ac{DO} change at each time step from advection, $DO_{adv}$. In practice, the simulated tidal signal was used to estimate $DO_{adv}$:
\begin{equation} \label{do_advp}
DO_{adv} \propto H
\end{equation}
\begin{equation} \label{do_adv}
DO_{adv} = 2\cdot a + a \cdot \frac{H- \min H}{\max H - \min H}
\end{equation}
where $a$ is analogous to $k$ in \cref{deltx} and is chosen as the transformation parameter to standardize change in \ac{DO} from tidal height change to desired units.  For example, $a = 1$ will convert $H$ to a scale that simulates changes in \ac{DO} from tidal advection that range from +/- 1 mg L$^{-1}$.  The final time series for observed \ac{DO} was the sum of biological \ac{DO} and advection \ac{DO} (\cref{fig:do_sim}).

@@

Multiple time series were simulated by varying the conditions in \cref{do_obs,do_bio,do_unc,do_obs_all,do_sin,do_unc_n,deltdo,deltx,do_advp,do_adv}. Specifically, the simulated data varied in the relative amount of noise in the measurement, relative amplitude of the diel \ac{DO} component, and degree of association of the tide with the \ac{DO} signal.  Three levels were evaluated for each variable: relative noise as 0, 1, and 2 standard deviations for both process and observation uncertainty, amplitude of diel biological \ac{DO} as 0, 1, and 2 mg L$^{-1}$, and \ac{DO} change from tidal advection as 0, 1, and 2 mg L$^{-1}$. A total of 81 time series were created based on the unique combinations of parameters (\cref{fig:sim_ex}).  Three half window widths for each variable (day, hour of day, and tide height) were evaluated: decimal time as 1, 3, and 6 days, time of day as 1, 3, and 6 hours, and tidal height as 0.25, 0.5, and 1 as a proportion of the observed value at the center of the window.  The window widths were chosen based on preliminary assessments that suggested variation in model performance was adequately captured by these values.  In total, 27 window width combinations were evaluated for each of 81 simulated time series, producing results for 2187 weighted regressions.

The detided or normalized values for each regression were compared to the simulated data to evaluate the ability of weighted regression to reproduce the biological \ac{DO} signals. Results were summarized using Pearson correlation coefficients and the \ac{RMSE} between the detided time series and the biological \ac{DO} time series as a known component of the observed.  Overall, the weighted regressions provided accurate results for the detided time series compared to the `true' biological time series regardless of the simulation parameters (\cref{tab:dtd_perf1}) or window widths (\cref{tab:dtd_perf2}).  Results for each simulation can be viewed using the link in the \hyperref[multi]{multimedia} section.  Mean correlation for all time series and window widths between the detided and biological values was \Sexpr{form_fun(perf_means['cor'])}, with values ranging from \Sexpr{form_fun(perf_rngs[1,'cor'])} to \Sexpr{form_fun(perf_rngs[2, 'cor'])}.  Mean error was \Sexpr{form_fun(perf_means['err'])}, with values ranging from \Sexpr{form_fun(perf_rngs[1,'err'])} to \Sexpr{form_fun(perf_rngs[2, 'err'])}.  Simulations with very poor performance (e.g., negative correlations) were those that had minimum widths for day windows and maximum widths for hour windows, or were those with the \ac{DO} signal composed entirely of noise from observation uncertainty. Conversely, simulations with detided time series that were identical to the true time series (e.g., correlation of one, \ac{RMSE} of zero) were those for which there was no biological or tidal influence.  While the latter examples do not represent real-world scenarios, they were included in the simulations to provide verification that the weighted regression provided reasonable results given extremes.

Characteristics of \ac{DO} time series that contributed to improved model performance were increasing amplitude of the diel \ac{DO} component ($DO_{die}$) and increasing process uncertainty ($e_{pro}$), whereas increasing observation uncertainty contributed to decreasing performance (\cref{tab:dtd_perf1,fig:err_surf1}).  Model performance was minimally influenced by magnitude of the tidal advection component ($DO_{adv}$), although performance decreased slightly with increasing tidal effects.  Increasing widths for day and tidal proportion windows contributed to increasing model performance, whereas the opposite was true for increasing hour windows (\cref{tab:dtd_perf2,err_surf2}).  Graphical summaries of model performance by simulation parameters (\cref{fig:err_surf1}) and half window widths (\cref{fig:err_surf2}) support the general trends described by \cref{tab:dtd_perf1,tab:dtd_perf2}.  Scale differences between \cref{fig:err_surf1} and \cref{fig:err_surf2} emphasize that model performance was more affected by characteristics of the \ac{DO} time series rather than the selected window widths.  For example, the range of correlation values comparing the effects of half window widths (averaged across all simulation parameters, \cref{fig:err_surf2}) were approximately half the range of correlations for comparing the effects of simulation parameters (averaged across all half window widths, \cref{fig:err_surf1}).

@@

\ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \ac{SWMP}.  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters attributed to both biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.

Continuous \ac{DO} time series and tidal height measurements at four sites from the \ac{SWMP} database \citepalias{CDMO14} were used to validate the detiding model with real data.  Monitoring data from January 1\textsuperscript{st} to December 31\textsuperscript{st} 2012 were obtained from a range of geographic locations (\cref{fig:case_map,tab:case_att}). Astronomical tidal heights were predicted for each site using harmonic regression applied to the sonde depth data (\texttt{oce} package in R, \citealt{Foreman89}, \citetalias{RDCT14}). Although, the depth data represent tidal height variation from both astronomical (i.e., gravitational effects) and meteorological (e.g., wind, precipitation inflows) sources, we isolated the former given that daily metabolism estimates were more likely to be affected by repeated diel cycling from normal tidal changes.  Each station was also chosen based on high correlations between \ac{DO} and tidal changes.  The four sites included the Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).  The stations are generally macrotidal semidiurnal or mixed semidiurnal and net heterotrophic on an annual basis.  Net heterotrophy (i.e., respiration exceeding production) is typical for shallow water systems at temperate latitudes \citep{Caffrey03}, although values in \cref{tab:case_att} are from observed \ac{DO} time series that are strongly influenced by tidal advection.

@@

The weighted regression method was applied to the time series for each station to obtain a detided \ac{DO} time series for estimating metabolism.  Half window widths of six days, one hour, and a tidal proportion of one half were chosen based on a balance between large and small window widths, although we recognize that the chosen widths are somewhat arbitrary and a more exhaustive evaluation should be conducted prior to using the results to inform management actions. Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}.  We hypothesized that metabolism esimates using the detided signal would contain less `anomalous' values than those from the observed \ac{DO} time series.  Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assume that physical processes are the dominant sources of these values.

Ecosystem metabolism was estimated from the \ac{DO} time series using the open-water technique \citep{Odum56} as described in \citet{Caffrey13}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange at the interface ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey13}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO}, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  The diffusion-corrected \ac{DO} flux estimates were averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of \ac{DO} change as the difference between observations at time $t$ and $t+1$.  Areal respiration rates were assumed constant during the night and substracted from daily gross production estimates to yield net ecosystem metabolism (\cref{tab:case_att}).

Detiding had significant effects on the correlations between tidal height changes, \ac{DO} time series, and metabolism estimates (\cref{tab:cor_res}).  Correlations of observed \ac{DO} time series with predicted tidal height were highly significant, with all sites indicating positive relationships, except Padilla Bay where tidal increases were associated with declines in \ac{DO}  concentration.  This suggests that seaward water masses were less anoxic than landward masses, with the opposite being true for Padilla Bay. The detided \ac{DO} time series had greatly reduced correlations with tidal height, although relationships were still significant after detiding likely because of the large sample size for each site (n $\approx$ 17500). Metabolism estimates before and after detiding were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Comparison of metabolic rates to tidal changes before and after detiding produced inconsistent results (\cref{tab:cor_res}).  Correlations for Elkhorn Slough and Sapelo Island showed consistent reductions in all three metabolims estimates after detiding.  Correlations for Padilla Bay and Rookery Bay were of opposite sign and greater magnitude after detiding for production and respiration, although net metabolism estimates had reduced correlations.

The percent of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for most sites after detiding (\cref{tab:case_res}).  Before detiding, anomalous values ranged from \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Padilla Bay) for production and \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Elkhorn Slough) for respiration as proportions of the daily estimates. Anomalous values were reduced to near zero for Rookery Bay and Sapelo Island, by approximately half for Padilla Bay (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for respiration), and only slightly reduced for Elkhorn Slough (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for respiration).  Metabolism estimates using detided \ac{DO} time series had decreased mean production and respiration for Elkhorn Slough, increased mean production and respiration for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Mean net ecosystem metabolism was unchanged for all sites.  Decreases in the standard erorr for all metabolism estimates (production, respiration, and net) were observed for all cases after detiding.  Graphical results for each case study can be viewed using the link in the \hyperref[multi]{multimedia} section.

>>>>

The ability of the weighted regression to detide the \ac{DO} signal was evaluated first using a simulation approach.  Observed \ac{DO} time series were created to represent the sum of variation from biological processes and physical effects related to tidal advection:  
\begin{equation} \label{do_obs}
DO_{obs} = DO_{bio} + DO_{adv}
\end{equation}
Biological \ac{DO} signals are inherently noisy \citep{Batt12} and can be further described as:
\begin{equation} \label{do_bio} 
DO_{bio} = DO_{die} + DO_{unc}
\end{equation} 
\begin{equation} \label{do_unc}
DO_{unc} = \epsilon_{obs} + \epsilon_{proc}
\end{equation}
where the biological \ac{DO} signal is the sum of diel variation plus uncertainty or noise.  Total uncertainty in the biological \ac{DO} signal is described as variation from observation and process uncertainty \citep{Hilborn97}.  Multiple time series at 30 minute time steps over 30 days were created following \cref{do_obs,do_bio,do_unc} such that observed \ac{DO} is generalized as the additive combination of four time series (\cref{fig:do_sim}):
\begin{equation} \label{do_obs_all}
DO_{obs} = DO_{adv} + DO_{die} + \epsilon_{obs} + \epsilon_{pro}
\end{equation}
Time series were created by varying the relative magnitudes of each of the four components of observed \ac{DO} to test the effectiveness of weighted regression under different scenarios.  The effects of air-sea gas exchange were not considered in the simulation given that methods are available for \textit{in situ} data to correct observed \ac{DO} for diffusion \citep[i.e., ][]{Thebault08}.

Each component of the simulated time series was created as follows.  First, the diel component, $DO_{die}$, was estimated using a sine/cosine function \citep{Cryer08}:
\begin{equation} \label{do_sin}
DO_{die} = \alpha + \beta\cos\left(2\pi ft + \Phi\right)
\end{equation}
such that the mean DO $\alpha$ was 8, amplitude $\beta$ was 1, $f$ was 1/48 to repeat on a 24 hour period every 30 minutes, $t$ was the time series vector and $\Phi$ was the x-axis origin set for an arbitrary sunrise at 630am.  The diel signal was increasing during the day and decreasing during the night for each 24 hour period and ranged from 7 to 9 mg L$^{-1}$.  Noise or uncertainty was added to the diel \ac{DO} signal to simulate natural variation in \ac{DO} throughout the time series (\cref{fig:do_sim}).  Total uncertainty was the sum of observation and process uncertainty for $n = 1440$ (30 minutes by 30 days) observations \citep{Hilborn97}, such that:
\begin{equation} \label{do_unc_n}
DO_{unc, n} = \epsilon_{obs, n} + \int_{t = 1}^{n} \epsilon_{pro, t}
\end{equation}
where observation and process uncertainty ($\epsilon_{obs}$, $\epsilon_{pro}$) were simulated as normally distributed random variables with mean zero and standard deviation varying from zero to an upper limit, described below.  To induce auto-correlation, process uncertainty was estimated as the cumulative sum of $n$ observations where the noise at time $t+1$ was equal to the noise at time $t$ plus additional variation drawn from the normal distribution.  The noise vector for process uncertainty was rescaled to constrain the variation within the bounds for standard deviation defined by the random variable. The total uncertainty, $DO_{unc}$, was added to the diel \ac{DO} time series to create the biological \ac{DO} time series (\cref{do_bio,fig:do_sim}).

A semidiurnal tidal series was simulated as a sine wave with a period of 12.5 hours to represent the principal lunar component \citep{Foreman89}.  The amplitude was set to 1 meter and centered  at 4 meters.  Initial assessments indicated that tide type (i.e., diurnal, semidiurnal, mixed) did not signifcantly affect the outcome of the results and a semidiurnal time series was used to reduce the total number of simulations.  The tidal time series was added to the biological \ac{DO} series to simulate \ac{DO} changes with advection, $DO_{adv}$ (\cref{do_obs_all,fig:do_sim}). Conceptually, this vector represents the rate of change in \ac{DO} as a function of horizontal water movement from tidal advection such that:
\begin{equation} \label{deltdo}
\frac{\delta DO_{adv}}{\delta t} = \frac{\delta DO}{\delta x} \cdot \frac{\delta x}{\delta t}
\end{equation}
\begin{equation} \label{deltx}
\frac{\delta x}{\delta t} = k \cdot \frac{\delta H}{\delta t}
\end{equation}
where the first derivative of the tidal time series, as change in height over time $\delta H / \delta t$, is multiplied by a constant $k$, to estimate horizontal tidal excursion over time, $\delta x / \delta t$.  The horizontal excursion is assumed to be associated with a horizontal \ac{DO} change, $\delta DO / \delta x$, such that the product of the two estimates the \ac{DO} change at each time step from advection, $DO_{adv}$. In practice, the simulated tidal signal was used to estimate $DO_{adv}$:
\begin{equation} \label{do_advp}
DO_{adv} \propto H
\end{equation}
\begin{equation} \label{do_adv}
DO_{adv} = 2\cdot a + a \cdot \frac{H- \min H}{\max H - \min H}
\end{equation}
where $a$ is analogous to $k$ in \cref{deltx} and is chosen as the transformation parameter to standardize change in \ac{DO} from tidal height change to desired units.  For example, $a = 1$ will convert $H$ to a scale that simulates changes in \ac{DO} from tidal advection that range from +/- 1 mg L$^{-1}$.  The final time series for observed \ac{DO} was the sum of biological \ac{DO} and advection \ac{DO} (\cref{do_obs,fig:do_sim}).

@@

Multiple time series were simulated by varying the conditions in \cref{do_obs,do_bio,do_unc,do_obs_all,do_sin,do_unc_n,deltdo,deltx,do_advp,do_adv}. Specifically, the simulated data varied in the relative amount of noise in the measurement ($e_{pro}$, $e_{obs}$), relative amplitude of the diel \ac{DO} component ($DO_{die}$), and degree of association of the tide with the \ac{DO} signal ($DO_{adv}$).  Three levels were evaluated for each variable: relative noise as 0, 1, and 2 standard deviations for both process and observation uncertainty, amplitude of diel biological \ac{DO} as 0, 1, and 2 mg L$^{-1}$, and \ac{DO} change from tidal advection as 0, 1, and 2 mg L$^{-1}$. A total of 81 time series were created based on the unique combinations of parameters (\cref{fig:sim_ex}).  Three half window widths for each variable (day, hour of day, and tide height) were evaluated for each simulation: decimal time as 1, 3, and 6 days, time of day as 1, 3, and 6 hours, and tidal height as 0.25, 0.5, and 1 as a proportion of the total range given the height at the center of the window.  The window widths were chosen based on preliminary assessments that suggested variation in model performance was adequately captured by these values.  In total, 27 window width combinations were evaluated for each of 81 simulated time series, producing results for 2187 weighted regressions.

The detided or normalized values for each regression were compared to the simulated data to evaluate the ability of weighted regression to reproduce the biological \ac{DO} signals. Results were summarized using Pearson correlation coefficients and the \ac{RMSE} between the detided time series and the biological \ac{DO} time series as a known component of the observed.  Overall, the weighted regressions provided accurate results for the detided time series compared to the `true' biological time series regardless of the simulation parameters (\cref{tab:dtd_perf1}) or window widths (\cref{tab:dtd_perf2},   results for each simulation can be viewed using the link in the \hyperref[multi]{multimedia} section).  Mean correlation for all time series and window widths between the detided and biological values was \Sexpr{form_fun(perf_means['cor'])}, with values ranging from \Sexpr{form_fun(perf_rngs[1,'cor'])} to \Sexpr{form_fun(perf_rngs[2, 'cor'])}.  Mean error was \Sexpr{form_fun(perf_means['err'])}, with values ranging from \Sexpr{form_fun(perf_rngs[1,'err'], 0, 0, 0)} to \Sexpr{form_fun(perf_rngs[2, 'err'])}.  Simulations with very poor performance (e.g., negative correlations) were those that had minimum widths for day windows and maximum widths for hour windows, or were those with the \ac{DO} signal composed entirely of noise from observation uncertainty. Conversely, simulations with detided time series that were identical to the true time series (e.g., correlation of one, \ac{RMSE} of zero) were those for which there was no biological or tidal influence.  Although these examples do not represent real-world scenarios, they were included in the simulations to provide verification that the weighted regression provided reasonable results given extremes.

Characteristics of \ac{DO} time series that contributed to improved model performance were increasing amplitude of the diel \ac{DO} component ($DO_{die}$) and increasing process uncertainty ($e_{pro}$), whereas increasing observation uncertainty contributed to decreased performance (\cref{tab:dtd_perf1,fig:err_surf1}).  Model performance was minimally influenced by magnitude of the tidal advection component ($DO_{adv}$), although performance decreased slightly with increasing tidal effects.  Increasing widths for day and tidal proportion windows contributed to increasing model performance, whereas the opposite was true for increasing hour windows (\cref{tab:dtd_perf2,fig:err_surf2}).  Graphical summaries of model performance by simulation parameters (\cref{fig:err_surf1}) and half window widths (\cref{fig:err_surf2}) support the general trends described by \cref{tab:dtd_perf1,tab:dtd_perf2}.  Scale differences between \cref{fig:err_surf1} and \cref{fig:err_surf2} emphasize that model performance was more affected by characteristics of the \ac{DO} time series rather than the selected window widths.  For example, the range of correlation values comparing the effects of half window widths (averaged across all simulation parameters, \cref{fig:err_surf2}) were approximately half the range of correlations for comparing the effects of simulation parameters (averaged across all half window widths, \cref{fig:err_surf1}).

@@

\ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \ac{SWMP}.  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.

Continuous \ac{DO} time series and tidal height measurements at four sites from the \ac{SWMP} database \citepalias{CDMO14} were used to validate the detiding model with real data.  Monitoring data from January 1\textsuperscript{st} to December 31\textsuperscript{st} 2012 were obtained from a range of geographic locations (\cref{fig:case_map,tab:case_att}). Astronomical tidal heights were predicted for each site using harmonic regression applied to the sonde depth data (\texttt{oce} package in R, \citealt{Foreman89}, \citetalias{RDCT14}). Although, the depth data represent tidal height variation from both astronomical (i.e., gravitational effects) and meteorological (e.g., wind, precipitation inflows) sources, we isolated the former given that daily metabolism estimates were more likely to be affected by repeated diel cycling from normal tidal changes.  Each station was also chosen based on highly significant correlations between \ac{DO} and tidal changes.  The four sites included the Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).  The stations are generally macrotidal semidiurnal or mixed semidiurnal and net heterotrophic on an annual basis (\cref{tab:case_att}).  Net heterotrophy (i.e., respiration exceeding production) is typical for shallow water systems at temperate latitudes \citep{Caffrey03}, although values in \cref{tab:case_att} are from observed \ac{DO} time series that are strongly influenced by tidal advection.

@@

The weighted regression method was applied to the time series for each station to obtain a detided \ac{DO} time series for estimating metabolism.  Half window widths of six days, one hour, and a tidal proportion of one half were chosen based on a balance between large and small window widths, although the chosen window widths are arbitrary and a more exhaustive evaluation should be conducted prior to using the results to inform management. Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Daily metabolism estimates before and after detiding were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}.  We hypothesized that metabolism esimates using the detided signal would contain less `anomalous' values than those from the observed \ac{DO} time series.  Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assume that physical processes are the dominant sources of these values given the tidal characteristics at each site.  Finally, means and standard errors of metabolism estimates were evaluated before and after detiding to determine if annual aggregations were significantly different.

Ecosystem metabolism was estimated from the \ac{DO} time series using the open-water technique \citep{Odum56} as described in \citet{Caffrey13}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange at the surface ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey13}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO}, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  The diffusion-corrected \ac{DO} flux estimates were averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of \ac{DO} change as the difference between observations at time $t$ and $t+1$ on an hourly scale.  Areal respiration rates were assumed constant during the night and substracted from daily gross production estimates to yield net ecosystem metabolism (\cref{tab:case_att}).

Detiding had significant effects on the correlations between tidal height changes, \ac{DO} time series, and metabolism estimates (\cref{tab:cor_res},  see the link in the \hyperref[multi]{multimedia} section for graphical results of each case study).  Correlations of observed \ac{DO} time series with predicted tidal height were highly significant, with all sites indicating positive relationships, except Padilla Bay where tidal increases were associated with declines in \ac{DO}  concentration.  This suggests that seaward water masses were less anoxic than landward masses, with the opposite being true for Padilla Bay. The detided \ac{DO} time series had greatly reduced correlations with tidal height, although relationships were still significant after detiding likely because of the large sample size for each site (n $\approx$ 17500). Comparison of metabolic rates to tidal changes before and after detiding produced inconsistent results (\cref{tab:cor_res}).  Correlations for Elkhorn Slough and Sapelo Island showed consistent reductions in all three metabolims estimates after detiding.  Correlations for Padilla Bay and Rookery Bay were of opposite sign and greater magnitude after detiding for production and respiration, although net metabolism estimates had reduced correlations.

The proportion of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for most sites after detiding (\cref{tab:case_res}).  Before detiding, anomalous values ranged from \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Pg', 'Value']))} (Padilla Bay) for production and \Sexpr{form_fun(min(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Rookery Bay) to \Sexpr{form_fun(max(anom_obs[anom_obs$Metab == 'Rt', 'Value']))} (Elkhorn Slough) for respiration. Anomalous values were reduced to near zero for Rookery Bay and Sapelo Island, by approximately half for Padilla Bay (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Padilla Bay', 'Value'])} for respiration), and only slightly reduced for Elkhorn Slough (\Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Pg' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for production, \Sexpr{form_fun(anom_dtd[anom_dtd$Metab == 'Rt' & anom_dtd$.id == 'Elkhorn Slough', 'Value'])} for respiration).  Metabolism estimates using detided \ac{DO} time series had decreased mean production and respiration for Elkhorn Slough, increased mean production and respiration for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Mean net ecosystem metabolism was unchanged for all sites.  Decreases in the standard erorr for all metabolism estimates (production, respiration, and net) were observed for all cases after detiding.
